### 2023-09-05 sections/2-methods.tex
fixed unfortunate typos

In this manuscript, we are primarily concerned with $k$-nearest neighbors in a finite-dimensional space. Given a dataset $\textbf{X} = \{x_1 \dots x_n\}$, we define a \emph{point} or \emph{datum} $x_i \in \textbf{X}$ as a singular observation (e.g., the genome of 
an organism, the neural embedding of a single image, or any entity on which we can define a \emph{distance function}).

We define a \emph{distance function} $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^+ \cup \{0\}$ as a function which, 
given two points $x, y \in \textbf{X}$, deterministically returns a non-negative real number. We require that the distance function 
be symmetric (i.e., $f(x, y) = f(y, x)$ for all $x, y \in \textbf{X}$) and that the distance between two points $x$ and $y$ be zero if and only if $x = y$. 
In addition to these constraints, if the distance function obeys
the triangle inequality, i.e., it is a \emph{distance metric}, then we guarantee that CAKES has perfect recall. 
Choice of distance function varies by type of data. For example, with electromagnetic spectra, we use both 
Euclidean (L2) and Cosine distances. With biological or string data, Hamming and Levenshtein distances are more appropriate.

Some of our algorithms for $\rho$- and $k$-nearest neighbors search rely on the \emph{local fractal dimension} at some point 
and length scale in the dataset. 
We define local fractal dimension (LFD) as: 
\begin{equation} \frac{\text{log}(\frac{|B_X(q, r_1)|}{|B_X(q, r_2)|})}{\text{log}(\frac{r_1}{r_2}) } \label{1} \end{equation}
where $B_X(q, r)$ is the set of points contained in a ball of radius $r$ 
centered at a point $q$ in the dataset $\textbf{X}$; in this example, we compute fractal dimension for some radius $r_1$ and a smaller radius $r_2 = \frac{r_1}{2}$.
We stress that this concept differs from the \emph{embedding dimension} of a dataset. To illustrate the difference,
consider SDSS's APOGEE dataset, wherein each datum is a nonnegative real-valued vector of length 8,575. Hence, the \emph{embedding dimension} of this dataset is 8,575. 
However, due to physical constraints (namely, the laws that govern stellar fusion and spectral emission lines), the data are constrained to a lower-dimensional 
manifold within the 8,575-dimensional embedding space. The \emph{local fractal dimension} is an approximation of the dimensionality of that lower-dimensional manifold in the ``vicinity'' of a given point.
This notion that high-dimensional data collected from constrained generating phenomena typically only occupy a low-dimensional manifold within their embedding space is known as the \emph{manifold hypothesis}~\cite{fefferman2016testing}.

@@

We define a \emph{cluster} as a set of points with a \emph{center}, a \emph{radius}, and an approximation of the \emph{local fractal dimension}.
The \emph{center} is the geometric median of a sample of points in the \emph{cluster}, and so it is a real data point. The \emph{radius} is the
maximum distance from a point in the cluster to its \emph{center}. We estimate \emph{local fractal dimension} at the cluster radius and half
the cluster radius. Each cluster (unless it is a leaf cluster) has two child clusters in much the same way that a node in
a binary tree has two child nodes. We define the \emph{metric entropy} $\mathcal{N}_{\hat{r_c}}(D)$ of a data set $D$ under a hierarchical clustering scheme as a refinement of [7], where
metric entropy for a given cluster radius $r_c$ was the number of clusters of radius $r_c$ needed to cover all data. Here, we use a hierarchical, divisive clustering 
approach, but with early stopping criteria; clusters which satisfy a user-specified stopping criterion (e.g. a specified cardinality or radius) are not further split. 
Since we frame the asymptotic complexity of $\rho$-NN search in terms of the number of leaf clusters, the \emph{metric entropy} is best thought of in terms of the number of
leaf clusters.

We start by building a divisive hierarchical clustering of the dataset with CLAM, using a 
similar recursive procedure as outlined in CHESS, but with the following 
improvements: better selection of poles for partitioning and depth-first dataset reordering 
(see Section ~\ref{subsubsec:methods:dataset-depth-first-reordering}).

@@

\subsubsection {Cluster Partitioning}

For a cluster $C$ with $|C|$ points, we begin by computing a 
random sample of $\sqrt{|C|}$ of its points, and computing pairwise distances 
between all points in the sample. Based on those pairwise distances, we compute the \emph{geometric median} of this sample; 
that is, the point which minimizes the sum of distances to all other points in the sample. This geometric median 
is $C$'s \emph{center}. The \emph{radius} of $C$ is the maximum distance from any point in $C$ to its center.
The point $l$ which is responsible for that radius (i.e., the furthest point from the center) is designated the \emph{left pole}. The point $r$ which is furthest
from $l$ is designated the \emph{right pole}. We then partition the cluster into a \emph{left child} and \emph{right child}, where the 
left child contains all points in the cluster which are closer to $l$ than to $r$, and the right child contains all 
points in the cluster which are closer to $r$ than to $l$. Without loss of generality, we assign to the left child 
those points which are equidistant from $l$ and $r$. Starting from a root-cluster containing the entire dataset, we 
repeat this procedure until each leaf contains only one datum or some other user-specified stopping criterion 
is met. This process is described in Algorithm \ref{alg:partition}.

\begin{algorithm} % enter the algorithm environment
\caption{Partition(\emph{C})} % give the algorithm a caption
\label{alg:partition} % and a label for \ref{} commands later in the document
\begin{algorithmic} % enter the algorithmic environment
    \STATE $m \Leftarrow \lfloor \sqrt{|C|} \rfloor$
    \STATE $seeds \Leftarrow m$ random points from $C.points$
    \STATE $c \Leftarrow$ geometric median of $seeds$
    \STATE $l \Leftarrow \argmax d(c,x) \ \forall \ x \in C.points$
    \STATE $r \Leftarrow \argmax d(l,x) \ \forall \ x \in C.points$
    \STATE $left \Leftarrow \{x | x \in C.points \land d(l,x) \le d(r,x)\}$
    \STATE $right \Leftarrow \{x | x \in C.points \land d(r,x) < d(l,x)\}$
    \IF{$|left| > 1$}
        \STATE Partition($left$)
    \ENDIF
    \IF{$|right| > 1$}
        \STATE Partition($right$)
    \ENDIF
\end{algorithmic}
\end{algorithm}

\subsubsection {Depth-First Reordering}
\label{subsubsec:methods:dataset-depth-first-reordering}

CAKES also improves upon CHESS by reordering the dataset in depth-first order of traversal of the tree.
In CHESS, each cluster stored a list of indices into the dataset, 
which we used during search to retrieve a cluster's points. 
Though this approach allowed us to retrieve the points in constant 
time, its memory cost was prohibitively high; since each cluster stores indices for each of its
points, for a dataset with cardinality $n$, we store a total of $n$ indices at each depth in the tree.
Assuming $\log{}n$ levels in the tree, this approach had $\mathcal{O}(n\log{}n)$ memory cost. 
[TODO: talk to Noah about how to phrase this, how to get at that balanced tree is the worst case]

One may improve this memory cost to $\mathcal{O}(n)$ by storing only these indices at 
the leaf cluster level.
This approach, however, introduces an $\mathcal{O}(n)$ time cost whenever we need to find the indices for 
a non-leaf cluster, since it requires a traversal of the subtree rooted at that cluster.

In this work, we introduce a new approach, wherein after building the cluster tree, we reorder the dataset 
so that points are stored in a depth-first order. Then, within each cluster, we need only 
store the cluster's cardinality and an \emph{offset} to access the cluster's points from the dataset. A left child has 
the same offset as its parent cluster, and the corresponding right child has an offset equal to the left child's offset
plus the left child's cardinality. With no additional memory cost nor time cost for retrieving points during search, 
dataset depth-first reordering offers significant improvements over the approach used in CHESS.

\subsubsection {Scaling Behavior of Cluster Radii}
\label{subsubsec:methods:guaranteed-decrease-in-cluster-radii}
While it may be tempting to assume that cluster radii decrease with each application of Partition (refer to Algorithm \ref{alg:partition}), unfortunately, this assumption is incorrect. 
Fortunately, we \emph{can} make some guarantees about the scaling behavior of cluster radii; in particular, we prove in this section that cluster radii will stop increasing after at most 
$d$ partitions, where $d$ is the fractal dimensionality of the dataset.

@@

In this case, Partition will select a maximally distant pair of points to use as poles; without loss of generality, 
say it chooses two points at the extrema one of the $d$ axes. 
After one partition, the maximum pairwise distance along that axis will be
bounded above by $R$. 
The next recursive Partition will select another of the $d$ axes, and once again, the distance along that axis will be bounded above by $R$
Thus, after at most $d$ partitions, the
maximum pairwise distance along each axis will be bounded above by $R$. 
The overall (i.e., not restricted to one axis) maximum pairwise distance 
will be bounded above by $R\sqrt{2}$ by, for example, two instances that lie at the extrema of different axes. 
See Figure~\ref{fig:methods:scaling_behavior} for an example.

@@

Note that, in practice, we never see a balanced clustering. Partition  produces unbalanced trees due to the varying density of the sampling 
in different regions of the manifold and the low dimensional ``shape'' of the manifold. Further, in practice, the cluster radii decrease by a factor much larger than 
$\frac{\sqrt{2}}{2}$ with every partition; the upper bound of $d$ partitions is seldom realized.

\begin{figure}[ht!]
    \centering
    \includegraphics[width=3.4in]{images/geometry/geometry.pdf}
    \caption{Scaling behavior of radii on a uniform-density disk, which represents the worst-case scenario for a two-dimensional distribution. 
    $R_0$ denotes the radius of the root cluster $C_0$, and $o_0$ denotes its center. 
    After one application of Algorithm~\ref{1}, we have $C_1$, with radius $R_1$ and center $o_1$.
    The right triangle formed by $o_0$, $o_1$, and $y_+$ in $C_1$ shows that $R_0 \leq R_1$.
    Hence, the radius of a child cluster can be larger than its parent. However, 
    after another application of Algorithm~\ref{1}, we have consumed all $d$ orthogonal axes, 
    as shown in $C_2$. 
    Now, clearly $R_2 \leq R_0$. In fact, for each subsequent application of Algorithm~\ref{1},
    the radius of the resulting cluster is bounded above by $\frac{\sqrt{2}R}{2}$.}
    \label{fig:methods:scaling_behavior}
\end{figure}

\subsubsection {Complexity}
\label{subsubsec:methods:clustering:complexity}

@@

\subsection {Sharding and Auto-Tuning}
\label{subsec:methods:index-building}
In this work, we also introduce a method of \emph{sharding} for datasets that cannot fit in memory. 
This approach also leverages the fact that $\rho$-nearest neighbors search at the radius of the $k$-th farthest neighbor 
is typically much faster than $k$-nearest neighbors search.
In addition to sharding, we perform some naive auto-tuning to select the optimal $k$-NN algorithm to be used with a given dataset.

With this approach, we begin by taking a random sample of no more than 10\% of the data. This sample 
is considered the first shard. 
The exact proportion of the dataset in this random sample depends on the dimensionality of the dataset and a user-specified memory limit. 
First, we build a cluster tree on this sample, and 
then consider all clusters at some user-specified depth in the tree, with a default depth of 10. Using these clusters' centers as queries, and a user-specified value of $k$, we record the time taken for $k$-NN search using each of the four algorithms described in Section~\ref{subsec:methods:knn-search}. We select the algorithm which had the lowest total search time over all the queries. 
We then split the rest of the dataset into shards based on the user-defined memory limit.

Sharded search then proceeds as follows. We use previously determined $k$ algorithm on the first shard. 
We use the distance from the $k$-th farthest neighbor to perform $\rho$-NN search on each subsequent shard. 
After searching each shard, we keep only the $k$-nearest neighbors so far, potentially decreasing the distance 
from the $k$-th farthest neighbor for search on the next shard.

Note that even though we select the optimal algorithm based on use with some user-specified value of $k$, we still allow search with any value of $k$.
Note also that the size of the shards is not necessarily consistent; it is often optimal to have the first shard be smaller than the others, since the first shard is the only one for which we perform $k$-nearest neighbors search. 
However, making this first shard too small introduces the risk of significantly overestimating the radius (i.e., the distance to the $k$-th neighbor), resulting in much slower $\rho$-nearest neighbors search on the subsequent shards.
Finding the precise optimal ratio between the size of the first shard and the size of the remaining shards, as well as computing the expected level of error between the distance from the $k$-th closest neighbor in the first shard and the actual $k$-th closest neighbor, are topics for future work. For datasets which \emph{do} fit in memory, we use the same approach to auto-tuning, but, since the entire dataset fits in memory, there is only one shard.

\subsection{The Search Problem}
\label{subsec:methods:general-search}
Having explained the clustering process, we can now pose the $k$- and $\rho$- nearest neighbors search problems.
Given a query $q$, along with a distance function $f$ defined on a dataset $\textbf{X}$, $k$-nearest neighbors search aims to find 
the set $S_q$ such that  $|S_q| = k$ and $\forall x \in \textbf{X} \setminus S_q$, $f(x, q) \leq \max\{f(y, q): y \in S_q \}$; that is,
for a given $k$, find the $k$ closest points to $q$ in $ \textbf{X}$.
We also have the $\rho$-nearest neighbors search problem, which aims to find the set $\{x \in \textbf{X}: f(q, x) \leq \rho \}$; that is, 
find all points in $\textbf{X}$ that are at most a distance $\rho$ from $q$.

Given a Cluster $C$, let $c$ be its center and $r$ be its radius. Our $\rho$- and $k$-NN algorithms make use of the following cluster 
properties:
\begin{itemize}
    \item $\delta = f(q, c)$ is the distance from the query to the cluster center $c$.
    \item $\delta_{max} = \delta + r$ is the distance from the query to the theoretically farthest possible instance in $C$.
    \item $\delta_{min} = \text{max}(0, \delta - r)$ is the distance from the query to the theoretically closest possible instance in $C$.
\end{itemize}

We define a \emph{singleton} as a cluster which either contains a single point (i.e., has cardinality 1) or which contains 
multiple instances of the same point (i.e., has cardinality greater than 1 but contains only one unique point). A singleton clearly 
has zero radius, and so $\delta_{max}  = \delta = \delta_{min}$ for these clusters. Hence, we sometimes overload the above 
notation to refer to distance from a query to a point; in these cases, we also have that $\delta_{max} = \delta = \delta_{min}$.

@@

We conduct $\rho$-nearest neighbors search as described in \cite{ishaq2019clustered}, but 
with the following improvement: when a cluster overlaps with the query ball, instead of  
always proceeding with both of its children, we proceed only with those children which 
also overlap with the query ball. This improvement is described in Algorithm \ref{alg:rnn}.

\begin{algorithm} 
    \caption{$\rho$-NN(\emph{clusters, query, r})} 
    \label{alg:rnn} 
    \begin{algorithmic}
        \REQUIRE $r \geq 0$
        \REQUIRE $clusters \neq \emptyset$
        \STATE $results \Leftarrow \emptyset$
        \IF{$clusters.left \land clusters.left \cap B_X(query, r) \neq \emptyset$}
            \IF{$clusters.left.\delta$ $\leq$ $r$ + $clusters.left.radius$}
                \STATE $\rho$-NN($clusters.left, query, r$)
            \ENDIF
        \ENDIF
        \IF{$clusters.right \land clusters.right \cap B_X(query, r) \neq \emptyset$}
            \IF{$clusters.right.\delta$ $\leq$ $r$ + $clusters.right.radius$} 
                \STATE $\rho$-NN($clusters.right, query, r$)
            \ENDIF
        \ENDIF
        \IF{$\neg$$clusters.left$ $\land \neg$$clusters.right$}
            \FOR{$p \in clusters.points$}
                \IF{$p.\delta \leq r$}
                    \STATE $results.push((p, r))$
                \ENDIF
            \ENDFOR
        \ENDIF
        \STATE Return $results$
    \end{algorithmic}
    \end{algorithm}

The asymptotic complexity of $\rho$-nearest neighbors is the same as in ~\cite{ishaq2019clustered}, namely:

\begin{gather}
    \mathcal{O}\Bigg(
    \underbrace{\log_2 \mathcal{N}_{\hat{r_c}}(D)}_{\textrm{metric entropy}} +
    \overbrace{\left|B_D(q,r)\right|}^{\textrm{output size}}
    \underbrace{\left(1+\frac{2\hat{r_c}}{r}\right)^d}_{\textrm{scaling factor}}
     \Bigg)
     \label{hierarchical-complexity}
\end{gather}

where $\hat{r_c}$ is the \textit{mean} cluster radius of leaf clusters, $\mathcal{N}_{\hat{r_c}}(D)$ is the metric entropy at that radius, $B_D(q,r)$ is a ball of data points around the query $q$ at search radius $r$, and $d$ is the fractal dimension around the query in a region of radius $r$.

@@

In this section, we present four novel algorithms for exact $k$-nearest neighbors search: $k$-NN by Repeated $\rho$-NN, Sieve Search, Sieve Search with Separate Centers, 
and Greedy Sieve. 
We use a preliminary auto-tuning function to predict the variant and shard size (see Section~\ref{subsec:methods:index-building}) which will perform 
best for a given query, dataset, and value of $k$. We then proceed with search using that variant on the first shard.

In these algorithms, we typically use $H$, for \emph{hits}, to refer to the data structure which stores the closest points to the query found so far and
$Q$ to refer to the data structure which stores the clusters (and, for Sieve and Sieve with Centers, points) which are still in contention for being one of the nearest neighbors.

@@

In this algorithm, we perform $\rho$-nearest neighbors search small search radius, repeatedly increasing the radius until $k$ neighbors
have been found.

Let $H$ be the set of nearest neighbors found thus far.
Search starts with a radius $m$ equal to the radius of the root cluster divided by
the cardinality of the dataset. We perform $\rho$-NN search with radius $m$. 
If no points are within a distance $m$ of the query, we increase the radius by a factor of
2 and perform $\rho$-NN search again, repeating until at least one point is found, i.e., 
until $|H| > 0$.

>>>>

In this manuscript, we are primarily concerned with $k$-nearest neighbors search in a finite-dimensional space.
Given a dataset $\textbf{X} = \{x_1 \dots x_n\}$, we define a \emph{point} or \emph{datum} $x_i \in \textbf{X}$ as a singular observation (e.g., the genome of 
an organism, the neural embedding of a single image, or any entity on which we can define a \emph{distance function}).

We define a \emph{distance function} $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^+ \ \cup \ \{0\}$ as a function which, 
given two points $x, y \in \textbf{X}$, deterministically returns a non-negative real number.
We require that the distance function be symmetric (i.e., $f(x, y) = f(y, x)$ for all $x, y \in \textbf{X}$) and that the distance between two points $x$ and $y$ be zero if and only if $x = y$. 
In addition to these constraints, if the distance function obeys the triangle inequality, i.e., it is a \emph{distance metric}, then we guarantee that CAKES has perfect recall. 
Choice of distance function varies by type of data.
For example, with electromagnetic spectra, one could both Euclidean (L2) and Cosine distances.
With biological or string data, Hamming and Levenshtein distances are more appropriate.

Some of our algorithms for $\rho$- and $k$-nearest neighbors search rely on the \emph{local fractal dimension} at some point and length scale in the dataset.
We define local fractal dimension (LFD) as:

\begin{equation}
    \frac{\text{log}(\frac{|B_X(q, r_1)|}{|B_X(q, r_2)|})}{\text{log}(\frac{r_1}{r_2}) }
    \label{eq:methods:lfd}
\end{equation}

where $B_X(q, r)$ is the set of points contained in a ball of radius $r$ centered at a point $q$ in the dataset $\textbf{X}$.
We stress that this concept differs from the \emph{embedding dimension} of a dataset.
To illustrate the difference, consider SDSS's APOGEE dataset, wherein each datum is a nonnegative real-valued vector of length 8,575.
Hence, the \emph{embedding dimension} of this dataset is 8,575. 
However, due to physical constraints (namely, the laws of physics that govern stellar fusion and spectral emission lines), the data are constrained to a lower-dimensional manifold within the 8,575-dimensional embedding space.
The \emph{local fractal dimension} is an approximation of the dimensionality of that lower-dimensional manifold in the ``vicinity'' of a given point.
This notion that high-dimensional data collected from constrained generating phenomena typically only occupy a low-dimensional manifold within their embedding space is known as the \emph{manifold hypothesis}~\cite{fefferman2016testing}.

@@

We define a \emph{cluster} as a set of points with a \emph{center}, a \emph{radius}, and an approximation of the \emph{local fractal dimension}.
The \emph{center} is the geometric median of a sample of points in the \emph{cluster}, and so it is a real data point.
The \emph{radius} is the maximum distance from any point in the cluster to the \emph{center}.
We estimate \emph{local fractal dimension} at the scale of the cluster radius and half the cluster radius.
Each cluster (unless it is a leaf cluster) has two child clusters in much the same way that a node in a binary tree has two child nodes.
We define the \emph{metric entropy} $\mathcal{N}_{\hat{r_c}}(D)$ of a data set $D$ under a hierarchical clustering scheme as a refinement of~\cite{yu2015entropy}, where metric entropy for a given cluster radius $r_c$ was the number of clusters of radius $r_c$ needed to cover all data.
Here, we use a hierarchical, divisive clustering approach, but with early stopping criteria; clusters which satisfy a user-specified stopping criterion (e.g. a specified cardinality or radius) are not further split.
Since we frame the asymptotic complexity of $\rho$-NN search in terms of the number of leaf clusters, the \emph{metric entropy} is best thought of in terms of the number of leaf clusters.

We start by building a divisive hierarchical clustering of the dataset with CLAM, using a similar recursive procedure as outlined in CHESS, but with the following improvements:
better selection of poles for partitioning and depth-first dataset reordering (see Section ~\ref{subsubsec:methods:clustering:dataset-depth-first-reordering}).

@@

\subsubsection {Cluster Partitioning}
\label{subsubsec:methods:cluster-partitioning}

For a cluster $C$ with $|C|$ points, we begin by taking a random sample of $\sqrt{|C|}$ of its points, and computing pairwise distances between all points in the sample.
Based on those pairwise distances, we compute the \emph{geometric median} of this sample; 
i.e., the point which minimizes the sum of distances to all other points in the sample.
This geometric median is $C$'s \emph{center}.
The \emph{radius} of $C$ is the maximum distance from any point in $C$ to its center.
The point $l$ which is responsible for that radius (i.e., the furthest point from the center) is designated the \emph{left pole}.
The point $r$ which is furthest from $l$ is designated the \emph{right pole}.
We then partition the cluster into a \emph{left child} and \emph{right child}, where the left child contains all points in the cluster which are closer to $l$ than to $r$, and the right child contains all points in the cluster which are closer to $r$ than to $l$.
Without loss of generality, we assign to the left child those points which are equidistant from $l$ and $r$.
Starting from a root-cluster containing the entire dataset, we repeat this procedure until each leaf contains only one datum or some other user-specified stopping criterion is met.
This process is described in Algorithm \ref{alg:methods:clustering:partition}.

\begin{algorithm} % enter the algorithm environment
\caption{Partition(\emph{C})} % give the algorithm a caption
\label{alg:methods:clustering:partition} % and a label for \ref{} commands later in the document
\begin{algorithmic} % enter the algorithmic environment
    \STATE $m \Leftarrow \lfloor \sqrt{|C|} \rfloor$
    \STATE $seeds \Leftarrow m$ random points from $C.points$
    \STATE $c \Leftarrow$ geometric median of $seeds$
    \STATE $l \Leftarrow \argmax d(c,x) \ \forall \ x \in C.points$
    \STATE $r \Leftarrow \argmax d(l,x) \ \forall \ x \in C.points$
    \STATE $left \Leftarrow \{x | x \in C.points \land d(l,x) \le d(r,x)\}$
    \STATE $right \Leftarrow \{x | x \in C.points \land d(r,x) < d(l,x)\}$
    \IF{$|left| > 1$}
        \STATE Partition($left$)
    \ENDIF
    \IF{$|right| > 1$}
        \STATE Partition($right$)
    \ENDIF
\end{algorithmic}
\end{algorithm}

\subsubsection {Depth-First Reordering}
\label{subsubsec:methods:clustering:dataset-depth-first-reordering}

CAKES also improves upon CHESS by reordering the dataset in depth-first order of traversal of the tree.
In CHESS, each cluster stored a list of indices into the dataset, which we used during search to retrieve a cluster's points. 
Though this approach allowed us to retrieve the points in constant time, its memory cost was prohibitively high;
since each cluster stores indices for each of its points and, for a dataset with cardinality $n$, we store a total of $n$ indices at each depth in the tree.
Assuming a balanced tree and thus $\mathcal{O}\log(n)$ depth, this approach had $\mathcal{O}(n\log(n))$ memory cost.

One may improve this memory cost to $\mathcal{O}(n)$ by only storing these indices at the leaf cluster level.
This approach, however, introduces an $\mathcal{O}(n\log(n))$ time cost whenever we need to find the indices for 
a non-leaf cluster, since it requires a traversal of the subtree rooted at that cluster.

In this work, we introduce a new approach, wherein after building the cluster tree, we reorder the dataset so that points are stored in a depth-first order.
Then, within each cluster, we need only store the cluster's cardinality and an \emph{offset} to access the cluster's points from the dataset.
The root cluster has an offset of zero and a cardinality equal to the number of points in the dataset.
A left child has the same offset as that of its parent, and the corresponding right child has an offset equal to the left child's offset plus the left child's cardinality.
With no additional memory cost nor time cost for retrieving points during search, depth-first reordering offers significant improvements over the approach used in CHESS.

\subsubsection {Scaling Behavior of Cluster Radii}
\label{subsubsec:methods:clustering:guaranteed-decrease-in-cluster-radii}

While it may be tempting to assume that cluster radii decrease with each application of Partition (refer to Algorithm \ref{alg:methods:clustering:partition}), unfortunately, this assumption is incorrect. 
Fortunately, we \emph{can} make some guarantees about the scaling behavior of cluster radii;
in particular, we prove in this section that cluster radii will stop increasing after at most $d$ partitions, where $d$ is the fractal dimensionality of the dataset.

@@

In this case, Partition will select a maximally distant pair of points to use as poles;
without loss of generality, say it chooses two points at the extrema one of the $d$ axes. 
After one partition, the maximum pairwise distance along that axis will be halved, i.e. bounded above by $R$.
The next recursive Partition will select another of the $d$ axes, and once again, the distance along that axis will be bounded above by $R$
Thus, after at most $d$ partitions, the maximum pairwise distance along each axis will be bounded above by $R$. 
The overall (i.e., not restricted to one axis) maximum pairwise distance 
will be bounded above by $R\sqrt{2}$ by, for example, two instances that lie at the extrema of different axes. 
See Figure~\ref{fig:methods:scaling-behavior} for an example.

@@

Note that, in practice, we never see a balanced clustering.
Partition produces unbalanced trees due to the varying density of the sampling in different regions of the manifold and the low dimensional ``shape'' of the manifold.
Further, in practice, the cluster radii decrease by a factor much larger than $\frac{\sqrt{2}}{2}$ with almost every partition;
the upper bound of $d$ partitions is seldom realized.

\begin{figure}[ht!]
    \centering
    \includegraphics[width=3.4in]{images/geometry/geometry.pdf}
    \caption{
        Scaling behavior of radii on a uniform-density disk, which represents the worst-case scenario for a two-dimensional distribution. 
        $R_0$ denotes the radius of the root cluster $C_0$, and $o_0$ denotes its center. 
        After one application of Algorithm~\ref{alg:methods:clustering:partition}, we have $C_1$, with radius $R_1$ and center $o_1$.
        The right triangle formed by $o_0$, $o_1$, and $y_+$ in $C_1$ shows that $R_0 < R_1$.
        Hence, the radius of a child cluster can be larger than its parent.
        However, after another application of Algorithm~\ref{alg:methods:clustering:partition}, we have consumed all $d$ orthogonal axes, 
        as shown in $C_2$.
        Now, clearly $R_2 < R_0$.
        In fact, for each subsequent application of Algorithm~\ref{alg:methods:clustering:partition}, the radius of the resulting cluster is bounded above by $\frac{\sqrt{2}R}{2}$.
    }
    \label{fig:methods:scaling-behavior}
\end{figure}

\subsubsection {Complexity}
\label{subsubsec:methods:clustering:clustering:complexity}

@@

\subsection {Sharding and Auto-Tuning}
\label{subsec:methods:sharding-and-auto-tuning}

In this work, we also introduce a method of \emph{sharding} for datasets that cannot fit in memory.
This approach also leverages the fact that $\rho$-nearest neighbors search at the radius of the $k$-th farthest neighbor is typically much faster than $k$-nearest neighbors search.
In addition to sharding, we perform some naive auto-tuning to select the optimal $k$-NN algorithm to be used with a given dataset.

With this approach, we begin by taking a random sample of no more than 10\% of the data.
This sample is considered the first shard.
The exact proportion of the dataset in this random sample depends on the dimensionality of the dataset and a user-specified memory limit. 
We then split the rest of the dataset into more shards based on the user-defined memory limit.
For each shard, we build a cluster tree using Algorithm~\ref{alg:methods:clustering:partition}.
From the first shard, we consider all clusters at some user-specified depth in the tree, with a default depth of 10.
Using these clusters' centers as queries, and a user-specified value of $k$, we record the time taken for $k$-NN search on the sample using each of the four algorithms described in Section~\ref{subsec:methods:knn-search}.
We select the algorithm which had the lowest total search time over all the queries.

Sharded search then proceeds as follows.
We use previously determined $k$-NN algorithm on the first shard. 
We use the distance to the $k$-th farthest neighbor to perform $\rho$-NN search on each subsequent shard. 
After searching each shard, we keep only the $k$-nearest neighbors so far, potentially decreasing the distance to the $k$-th farthest neighbor for search on the next shard.

Note that even though we select the optimal algorithm based on use with some user-specified value of $k$, we still allow search with any value of $k$.
Note also that the size of the shards is not necessarily consistent;
it is often optimal to have the first shard be smaller than the others, since the first shard is the only one for which we perform $k$-nearest neighbors search. 
However, making this first shard too small introduces the risk of significantly overestimating the radius (i.e., the distance to the $k$-th neighbor), resulting in much slower $\rho$-nearest neighbors search on the subsequent shards.
Finding the optimal ratio between the size of the first shard and the sizes of the remaining shards, as well as computing the expected level of error between the distance from the $k$-th closest neighbor in the first shard and the actual $k$-th closest neighbor, are topics for future work.
For datasets which \emph{do} fit in memory, we use the same approach to auto-tuning, but, since the entire dataset fits in memory, there is only one shard.

\subsection{The Search Problem}
\label{subsec:methods:the-search-problem}
Having explained the clustering process, we can now pose the $k$-NN and $\rho$-NN search problems.
Given a query $q$, along with a distance function $f$ defined on a dataset $\textbf{X}$, $k$-nearest neighbors search aims to find 
the set $S_q$ such that  $|S_q| = k$ and $\forall x \in \textbf{X} \ \setminus \ S_q$, $f(x, q) > \max\{f(y, q): y \in S_q \}$;
that is, for a given $k$, find the $k$ closest points to $q$ in $ \textbf{X}$.
We also have the $\rho$-nearest neighbors search problem, which aims to find the set $\{x \in \textbf{X}: f(q, x) \leq \rho \}$;
that is, find all points in $\textbf{X}$ that are at most a distance $\rho$ from $q$.

Given a Cluster $C$, let $c$ be its center and $r$ be its radius. Our $\rho$- and $k$-NN algorithms make use of the following properties:

\begin{itemize}
    \item $\delta = f(q, c)$ is the distance from the query to the cluster center $c$.
    \item $\delta_{max} = \delta + r$ is the distance from the query to the theoretically farthest possible instance in $C$.
    \item $\delta_{min} = \text{max}(0, \delta - r)$ is the distance from the query to the theoretically closest possible instance in $C$.
\end{itemize}

We define a \emph{singleton} as a cluster which either contains a single point (i.e., has cardinality 1) or which contains 
duplicates of the same point (i.e., has cardinality greater than 1 but contains only one unique point).
A singleton clearly has zero radius, and so $\delta_{max} = \delta = \delta_{min}$ for these clusters.
Hence, we sometimes overload the above notation to also refer to distance from a query to a point;
in these cases, we also have that $\delta_{max} = \delta = \delta_{min}$.

@@

We conduct $\rho$-nearest neighbors search as described in \cite{ishaq2019clustered}, but with the following improvement:
when a cluster overlaps with the query ball, instead of  always proceeding with both of its children, we proceed only with those children which also overlap with the query ball.
This improvement is described in Algorithm \ref{alg:methods:rnn-search}.

\begin{algorithm} 
    \caption{$\rho$-NN(\emph{cluster, query, r})} 
    \label{alg:methods:rnn-search} 
    \begin{algorithmic}
        \REQUIRE $r \geq 0$
        \REQUIRE $cluster \neq \emptyset$
        \STATE $results \Leftarrow \emptyset$
        \IF{$cluster.left \cap B_X(query, r) \neq \emptyset$}
            \IF{$cluster.left.\delta$ $\leq$ $r$ + $cluster.left.radius$}
                \STATE $\rho$-NN($cluster.left, query, r$)
            \ENDIF
        \ENDIF
        \IF{$cluster.right \cap B_X(query, r) \neq \emptyset$}
            \IF{$cluster.right.\delta$ $\leq$ $r$ + $cluster.right.radius$} 
                \STATE $\rho$-NN($cluster.right, query, r$)
            \ENDIF
        \ENDIF
        \IF{$cluster.isLeaf$}
            \FOR{$p \in cluster.points$}
                \IF{$p.\delta \leq r$}
                    \STATE $results.push((p, r))$
                \ENDIF
            \ENDFOR
        \ENDIF
        \STATE Return $results$
    \end{algorithmic}
\end{algorithm}

The asymptotic complexity of $\rho$-nearest neighbors is the same as in~\cite{ishaq2019clustered}, namely:

\begin{gather}
    \mathcal{O}\Bigg(
    \underbrace{\log_2 \mathcal{N}_{\hat{r_c}}(X)}_{\textrm{metric entropy}} +
    \overbrace{\left|B_X(q,r)\right|}^{\textrm{output size}}
    \underbrace{\left(1+\frac{2\hat{r_c}}{r}\right)^d}_{\textrm{scaling factor}}\Bigg)
    \label{hierarchical-complexity}
\end{gather}

where $\hat{r_c}$ is the \textit{mean} cluster radius of leaf clusters, $\mathcal{N}_{\hat{r_c}}(X)$ is the metric entropy at that radius, $B_X(q,r)$ is a ball of data points around the query $q$ at search radius $r$, and $d$ is the fractal dimension around the query in a region of radius $r$.

@@

In this section, we present four novel algorithms for exact $k$-nearest neighbors search: $k$-NN by Repeated $\rho$-NN, Sieve Search, Sieve Search with Separate Centers, and Greedy Sieve. 
We use a preliminary auto-tuning function to predict the variant (see Section~\ref{subsec:methods:sharding-and-auto-tuning}) which will perform 
best for a given dataset and value of $k$.
We then proceed with search using that variant on the first shard.

In these algorithms, we typically use $H$, for \emph{hits}, to refer to the data structure which stores the closest points to the query found so far and
$Q$ to refer to the data structure which stores the clusters and points which are still in contention for being one of the nearest neighbors.

@@

In this algorithm, we perform $\rho$-nearest neighbors search starting with a small search radius, and repeatedly increasing the radius until we find $k$ neighbors.

Let $H$ be the set of nearest neighbors found thus far.
Search starts with a radius $m$ equal to the radius of the root cluster divided by the cardinality of the dataset.
We perform $\rho$-NN search with radius $m$. 
If no points are within a distance $m$ of the query, we increase the radius by a factor of 2 and perform $\rho$-NN search again, repeating until at least one point is found.

Once $|H| > 0$, we continue to perform $\rho$-NN search, but instead of increasing the radius by a factor of 2 on each iteration, we increase it by a factor determined by the local fractal dimension in the region of the dataset around the query ball. 
In particular, we increase the radius by a factor of

\begin{equation}
    \min \left(2, \left({\frac{k}{|H|}}\right)^{\frac{1}{\mu}}\right)
    \label{eq:repeated-rnn-factor}
\end{equation}
